Deployable Probabilistic Programming

### 2019-08-12 paper/paper.tex
more introductory examples

@@

  Probabilistic programming \cite{GMR+08,MSP14,WVM14,GS15}
  represents statistical models as programs written in an
  otherwise general programming language that provides syntax
  for the definition and conditioning of random variables.
  Inference can be performed on probabilistic programs to
  obtain the posterior distribution or point estimates of the
  variables. Inference algorithms are provided by the
  probabilistic programming framework, and each algorithm is
  usually applicable to a wide class of probabilistic programs
  in a black-box manner. The algorithms include
  Metropolis-Hastings\cite{WSG11,MSP14,YHG14}, Hamiltonian
  Monte Carlo~\cite{Stan17}, expectation
  propagation~\cite{MWG+10}, extensions of Sequential Monte
  Carlo~\cite{WVM14,MYM+15,PWD+14,RNL+2016}, variational
  inference~\cite{WW13,KTR+17}, gradient-based
  optimization~\cite{Stan17,BCJ+19}, and others.

\begin{figure}
\begin{lstlisting}[language=WebPPL]
var breastCancer = flip(0.01)
var benignCyst = flip(0.2)
var positiveMammogram = (breastCancer && flip(0.8))
                      || (benignCyst && flip(0.5))
condition(positiveMammogram)
return breastCancer
\end{lstlisting}
	\caption{A probabilistic program in WebPPL~\cite{GS15}:
	probability of breast cancer given a positive mammogram.}
	\label{fig:mammo}
\end{figure}
  Consider, for example, the probabilistic program in
  Figure~\ref{fig:mammo}. The program models dependency between
  a positive mammogram and possible causes --- breast cancer
  or a benign cyst.  The program defines three Bernoulli
  random variables  --- \lstinline{breastCancer},
  \lstinline{benignCyst}, \lstinline{positiveMammogram} ---
  and conditions on one of the random variables,
  \lstinline{positiveMammogram}, being true. Inference on the
  program returns the probability of breast cancer given a
  positive mammogram.

@@

Let us illustrate the use of Infergo on a basic example serving
as a probabilistic programming equivalent of the "Hello, World!"
program --- inferring parameters of a unidimensional Normal
distribution given a set of data samples. The two parameters of
a Normal distribution are the mean $\mu$ and the standard
deviation $\sigma$. The solution can of course be obtained
analytically, but the organization of implementation and
inference possesses all of the elements of probabilistic
programming for larger, more complicated problems, for which a
closed form solution does not exist or difficult to derive.

@@

\begin{lstlisting}
hmc := &infer.HMC{
  Eps: 0.1,
}
samples := make(chan []float64)
hmc.Sample(m, x, samples)
for i := 0; i != 5000; i++ {
  sample = <-samples
}
hmc.Stop()
\end{lstlisting}

@@

Any user-defined function of appropriate kind can be used as an
elemental inside model methods. The gradient must be registered
for the function using a call to
\begin{lstlisting}
func RegisterElemental(
  f interface{},
  g func(value float64,
         params ...float64) []float64)
\end{lstlisting} For
example, the call
\begin{lstlisting}
RegisterElemental(Sigm,
  func(value float64,
       _ ...float64) []float64 {
    return []float64{value*(1-value)}
  })
\end{lstlisting}
registers the gradient for the \textit{sigmoid} function:
\begin{lstlisting}
func Sigm(x float64) float64 {
	return 1. / (1. + math.Exp(-x))
}
\end{lstlisting}

@@

\appendix

>>>>

\lstdefinelanguage{Anglican}%
{morekeywords={*,*1,*2,*3,*agent*,*allow-unresolved-vars*,*assert*,*clojure-version*,*command-line-args*,%
		*compile-files*,*compile-path*,*e,*err*,*file*,*flush-on-newline*,*in*,*macro-meta*,%
		*math-context*,*ns*,*out*,*print-dup*,*print-length*,*print-level*,*print-meta*,*print-readably*,%
		*read-eval*,*source-path*,*use-context-classloader*,*warn-on-reflection*,+,-,->,->>,..,/,:else,%
		<,<=,=,==,>,>=,@,accessor,aclone,add-classpath,add-watch,agent,agent-errors,aget,alength,alias,%
		all-ns,alter,alter-meta!,alter-var-root,amap,ancestors,and,apply,areduce,array-map,aset,%
		aset-boolean,aset-byte,aset-char,aset-double,aset-float,aset-int,aset-long,aset-short,assert,%
		assoc,assoc!,assoc-in,associative?,atom,await,await-for,await1,bases,bean,bigdec,bigint,binding,%
		bit-and,bit-and-not,bit-clear,bit-flip,bit-not,bit-or,bit-set,bit-shift-left,bit-shift-right,%
		bit-test,bit-xor,boolean,boolean-array,booleans,bound-fn,bound-fn*,butlast,byte,byte-array,%
		bytes,case,cast,char,char-array,char-escape-string,char-name-string,char?,chars,chunk,chunk-append,%
		chunk-buffer,chunk-cons,chunk-first,chunk-next,chunk-rest,chunked-seq?,class,class?,%
		clear-agent-errors,clojure-version,coll?,comment,commute,comp,comparator,compare,compare-and-set!,%
		compile,complement,concat,cond,condp,conj,conj!,cons,constantly,construct-proxy,contains?,count,%
		counted?,create-ns,create-struct,cycle,dec,decimal?,declare,def,defdist,definline,defm,defmacro,defmethod,%
		defmulti,defn,defn-,defonce,defquery,defproc,defprotocol,defrecord,defstruct,deftype,delay,delay?,deliver,deref,derive,%
		descendants,destructure,disj,disj!,dissoc,dissoc!,distinct,distinct?,do,do-template,doall,doc,%
		dorun,doseq,dosync,dotimes,doto,double,double-array,doubles,drop,drop-last,drop-while,empty,empty?,%
		ensure,enumeration-seq,eval,even?,every?,false,false?,ffirst,file-seq,filter,finally,find,find-doc,%
		find-ns,find-var,first,float,float-array,float?,floats,flush,fm,fn,fn?,fnext,for,force,format,future,%
		future-call,future-cancel,future-cancelled?,future-done?,future?,gen-class,gen-interface,gensym,%
		get,get-in,get-method,get-proxy-class,get-thread-bindings,get-validator,hash,hash-map,hash-set,%
		identical?,identity,if,if-let,if-not,ifn?,import,in-ns,inc,init-proxy,instance?,int,int-array,%
		integer?,interleave,intern,interpose,into,into-array,ints,io!,isa?,iterate,iterator-seq,juxt,%
		key,keys,keyword,keyword?,last,lazy-cat,lazy-seq,let,letfn,line-seq,list,list*,list?,load,load-file,%
		load-reader,load-string,loaded-libs,locking,long,long-array,longs,loop,macroexpand,macroexpand-1,%
		make-array,make-hierarchy,map,map?,mapcat,max,max-key,mem,memfn,memoize,merge,merge-with,meta,%
		method-sig,methods,min,min-key,mod,monitor-enter,monitor-exit,name,namespace,neg?,new,newline,%
		next,nfirst,nil,nil?,nnext,not,not-any?,not-empty,not-every?,not=,ns,ns-aliases,ns-imports,%
		ns-interns,ns-map,ns-name,ns-publics,ns-refers,ns-resolve,ns-unalias,ns-unmap,nth,nthnext,num,%
		number?,observe,odd?,or,parents,partial,partition,pcalls,peek,persistent!,pmap,pop,pop!,pop-thread-bindings,%
		pos?,pr,pr-str,predict,prefer-method,prefers,primitives-classnames,print,print-ctor,print-doc,print-dup,%
		print-method,print-namespace-doc,print-simple,print-special-doc,print-str,printf,println,println-str,%
		prn,prn-str,promise,proxy,proxy-call-with-super,proxy-mappings,proxy-name,proxy-super,%
		push-thread-bindings,pvalues,query,quot,rand,rand-int,range,ratio?,rational?,rationalize,re-find,%
		re-groups,re-matcher,re-matches,re-pattern,re-seq,read,read-line,read-string,recur,reduce,ref,%
		ref-history-count,ref-max-history,ref-min-history,ref-set,refer,refer-clojure,reify,%
		release-pending-sends,rem,remove,remove-method,remove-ns,remove-watch,repeat,repeatedly,%
		replace,replicate,require,reset!,reset-meta!,resolve,rest,resultset-seq,retrieve,reverse,reversible?,%
		rseq,rsubseq,sample,second,select-keys,send,send-off,seq,seq?,seque,sequence,sequential?,set,set!,%
		set-validator!,set?,short,short-array,shorts,shutdown-agents,slurp,some,sort,sort-by,sorted-map,%
		sorted-map-by,sorted-set,sorted-set-by,sorted?,special-form-anchor,special-symbol?,split-at,%
		split-with,store,str,stream?,string?,struct,struct-map,subs,subseq,subvec,supers,swap!,symbol,symbol?,%
		sync,syntax-symbol-anchor,take,take-last,take-nth,take-while,test,the-ns,throw,time,to-array,%
		to-array-2d,trampoline,transient,tree-seq,true,true?,try,type,unchecked-add,unchecked-dec,%
		unchecked-divide,unchecked-inc,unchecked-multiply,unchecked-negate,unchecked-remainder,%
		unchecked-subtract,underive,unquote,unquote-splicing,update-in,update-proxy,use,val,vals,%
		var,var-get,var-set,var?,vary-meta,vec,vector,vector?,when,when-first,when-let,when-not,%
		while,with-bindings,with-bindings*,with-in-str,with-loading-context,with-local-vars,%
		with-meta,with-open,with-out-str,with-precision,xml-seq,zero?,zipmap
	   sensitive,% ???
	      alsodigit=-,%
		     morecomment=[l];,%
			    morestring=[b]"%
				  }[keywords,comments,strings]%

@@

  Probabilistic programming \cite{GMR+08,MSP14,WVM14,GS15}
  represents statistical models as programs written in an
  otherwise general programming language that provides syntax
  for the definition and conditioning of random variables.
  Inference can be performed on probabilistic programs to
  obtain the posterior distribution or point estimates of the
  variables. Inference algorithms are provided by the
  probabilistic programming framework, and each algorithm is
  usually applicable to a wide class of probabilistic programs
  in a black-box manner. The algorithms include
  Metropolis-Hastings\cite{WSG11,MSP14,YHG14}, Hamiltonian
  Monte Carlo~\cite{Stan17}, expectation
  propagation~\cite{MWG+10}, extensions of Sequential Monte
  Carlo~\cite{WVM14,MYM+15,PWD+14,RNL+2016,MS18}, variational
  inference~\cite{WW13,KTR+17}, gradient-based
  optimization~\cite{Stan17,BCJ+19}, and others.

\begin{figure}
\begin{lstlisting}[language=WebPPL]
var breastCancer = flip(0.01)
var benignCyst = flip(0.2)
var positiveMammogram = (breastCancer && flip(0.8))
                      || (benignCyst && flip(0.5))
condition(positiveMammogram)
return breastCancer
\end{lstlisting}
  \caption{A probabilistic program in WebPPL~\cite{GS15}:
  probability of breast cancer given a positive mammogram.}
  \label{fig:mammo-webppl}
\end{figure}
  For example, the probabilistic program in
  Figure~\ref{fig:mammo-webppl} models dependency between
  a positive mammogram and possible causes --- breast cancer
  or a benign cyst.  The program defines three Bernoulli
  random variables  --- \lstinline{breastCancer},
  \lstinline{benignCyst}, \lstinline{positiveMammogram} ---
  and conditions on one of the random variables,
  \lstinline{positiveMammogram}, being true. Inference on the
  program returns the probability of breast cancer given a
  positive mammogram.

  \begin{figure}
    \begin{lstlisting}[language=Anglican]
(defquery linearRegression [xs ys x*]
  (let [m     (sample (normal 0 2))
        b     (sample (normal 0 2))
        sigma (sample (gamma 1 1))
        f     (fn [x] (+ (* m x) b))]
   (loop [data (map #(-> [%1 %2]) xs ys)]
    (when (seq data)
     (let [[[x y] & data] data] 
       (observe (normal (f x) sigma) y)
       (recur data))))
   (predict (f x*))))
    \end{lstlisting}
    \caption{A probabilistic program in Anglican~\cite{TMY+16}:
    linear regression.}
    \label{fig:linreg-anglican}
  \end{figure}
  Probabilistic programs may contain loops, conditional
  statements, recursion, and operate on built-in and user
  defined data structures. The probabilistic program in
  Figure~\ref{fig:linreg-anglican} implements linear regression
  (approximating dependency between the input and the output by
  a linear function) and contains loops
  (\lstinline[language=Anglican]{loop/recur}), conditional
  statements (\lstinline[language=Anglican]{when}), and
  user-defined functions (\lstinline[language=Anglican]{fn});
  the data is passed to the program as vectors of real numbers
  (\lstinline[language=Anglican]{xs}, \lstinline[language=Anglican]{ys}).

@@

Let us illustrate the use of Infergo on a basic example serving
as a probabilistic programming equivalent of the "Hello, World!"
program --- inferring parameters of a unidimensional Normal
distribution given a set of data samples, or observations. The
two parameters of a Normal distribution are the mean $\mu$ and
the standard deviation $\sigma$. The solution can of course be
obtained analytically, but the organization of implementation
and inference possesses all of the elements of probabilistic
programming for larger, more complex problems, for which a
closed form solution does not exist or is difficult to derive.

@@

\begin{lstlisting}
hmc := &infer.HMC{
  Eps: 0.1,
}
samples := make(chan []float64)
hmc.Sample(m, x, samples)
for i := 0; i != 5000; i++ {
  sample = <-samples
  // ...
}
hmc.Stop()
\end{lstlisting}

@@

Any user-defined function of appropriate kind can be used as an
elemental inside model methods. The gradient must be registered
for the function using a call to
\begin{lstlisting}
func RegisterElemental(
  f interface{},
  g func(value float64,
         params ...float64) []float64)
\end{lstlisting}
\lstinline{RegisterElemental} receives both the parameters and
the value of the function for efficient computation re-use, and
the gradient can be computed using either the parameters, the
value, or a combination of both~\cite{GW08}. For example, the
call
\begin{lstlisting}
RegisterElemental(Sigm,
  func(value float64,
       _ ...float64) []float64 {
    return []float64{value*(1-value)}
  })
\end{lstlisting}
registers the gradient for the \textit{sigmoid} function
\begin{lstlisting}
func Sigm(x float64) float64 {
  return 1. / (1. + math.Exp(-x))
}
\end{lstlisting}
and re-uses the function value to compute the gradient
efficiently, avoiding an expensive call to \lstinline{math.Exp}.

@@

\clearpage
\appendix
